---
title: "NFL Deliverable - Descriptive Analytics"
author: "Will and AJ"
date: "2026-04-18"
output: html_document
---
# Introduction
Our question of interest is the what parameters are most responsoible for determining if a NFL team will have a winnin season or not. For the purposes of this study, a winning season is defined as a season in which a team finishes with at least as many wins as losses. Therefore, seasons with records such as 8–8 or 8–8–1 are classified as winning seasons. We created a binary variable called winning_record, where a value of 1 indicates a winning season and a value of 0 indicates a losing season.

# Data
The dataset used in this analysis contains team-level statistics for NFL seasons, including offensive and defensive performance measures such as passing efficiency, rushing efficiency, scoring statistics, and turnovers. These data were obtained from an online sports analytics database containing yearly team performance statistics.
Although the dataset includes observations beginning in 1999, we restricted the analysis to seasons beginning in 2006. The earlier seasons contained missing or incomplete values for several variables that were important for our analysis. Because incomplete data can bias regression estimates and reduce model reliability, we filtered the dataset to include only seasons from 2006 onward, ensuring that all relevant variables were consistently recorded. This filtering step improved data quality and ensured that comparisons across teams and seasons were based on complete information.


```{r setup, include=FALSE}
library(tidyverse)
library(GGally)
library(MASS)
nfl_data <- read.csv("https://data.scorenetwork.org/data/nfl-team-statistics.csv")


nfl_data$win_diff <- nfl_data$wins - nfl_data$losses

nfl_data$winning_record <- ifelse(nfl_data$win_diff >= 0, 1, 0)


nfl_data <- nfl_data %>%
  filter(season >= 2006)


str(nfl_data)

```


Through our literature review it was previously theorized that turnovers were of large importance. We decided to put this to the test with a simple linear regression model.

```{r}
nfl_data$turnovers <- nfl_data$offense_n_interceptions +
                      nfl_data$offense_n_fumbles_lost_pass +
                      nfl_data$offense_n_fumbles_lost_run

turnover_model <- lm(wins ~ offense_ave_yards_gained_pass +
          offense_ave_yards_gained_run +
          turnovers,
   data = nfl_data)
summary(turnover_model)
```
