---
title: "NFL Deliverable - Descriptive Analytics"
author: "Will and AJ"
date: "2026-04-18"
output: html_document
---
# Introduction
Our question of interest is the what parameters are most responsoible for determining if a NFL team will have a winnin season or not. For the purposes of our models we have also called a season with an equal number of wins and losses i.e. 8-8-1 a winning season. 
```{r setup, include=FALSE}
library(tidyverse)
library(GGally)
library(MASS)
nfl_data <- read.csv("https://data.scorenetwork.org/data/nfl-team-statistics.csv")


nfl_data$win_diff <- nfl_data$wins - nfl_data$losses

nfl_data$winning_record <- ifelse(nfl_data$win_diff >= 0, 1, 0)

str(nfl_data)

```


Through our literature review it was previously theorized that turnovers were of large importance. We decided to put this to the test with a simple linear regression model.

```{r}
nfl_data$turnovers <- nfl_data$offense_n_interceptions +
                      nfl_data$offense_n_fumbles_lost_pass +
                      nfl_data$offense_n_fumbles_lost_run

turnover_model <- lm(wins ~ offense_ave_yards_gained_pass +
          offense_ave_yards_gained_run +
          turnovers,
   data = nfl_data)
summary(turnover_model)
```
